### 2014-10-28 contract-language.tex

Quantifier-free propositions in our contract language are conjunctions,
disjunctions and implications of predicates expressing relations between pairs
of effect variables. The syntactic class of relations is seeded with primitive
$\visZ$, $\soZ$, and $\sameobjZ$ relations, and also admits derived relations
that are expressible as union, intersection, or transitive
closure\footnote{Strictly speaking, $R^{+}$ is not the transitive closure of
$R$, as transitive closure is not expressible in FOL. Instead, $R^{+}$ in our
language denotes \emph{a} superset of transitive closure of $R$. Formally,
$R^{+}$ is any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow R'(x,z)$} of
primitive relations. Commonly used derived relations are the \emph{same object
session order} ($\small \sooZ = \soZ ~\cap~ \sameobjZ$), the \emph{happens
before order} ($\small \hbZ = (\soZ ~\cup~ \visZ)^+$), and the per-object
happens-before order ($\small \hboZ = (\sooZ ~\cup~ \visZ)^+$).

@@

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}.  Completeness of the logic~\cite{completeness} allows us
to claim that $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st}
\Rightarrow \cv_{op}$. Due to the decidability of our contract language, this
implication check is automatically discharged with the help of a theorem
prover. In this case, we say that $\cv_{op}$ is \emph{weaker than} $\cv_{st}$
(written $\cv_{op} \le \cv_{st}$).

@@

\vspace{-1em}
\begin{smathpar}
\ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}
\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
	contracts are required to see a causally consistent snapshot of the object
	state, including the actions performed on the same session.  The latter
	requirement entails that if two operations $o_1$ and $o_2$ from the same
	session are applied to two different replicas $r_1$ and $r_2$, the second
	operation cannot be discharged until the effect of $o_1$ is merged with $o_2$
	in both $r_1$ and $r_2$.  \cf{getBalance} operation is classified as a
	causally consistent operation. We assume that causality is only tracked
	through operations on the same object; two operations in the same session but
	on different objects are considered causally unrelated under this definition.
	Stores typically avoid tracking causality across objects to mitigate
	overheads when causality tracking is unnecessary.  The corresponding store
	semantics is captured by the contract $\ccc$ presented in
	Section~\ref{sec:store_sem}.

@@

\noindent Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves
totally ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$.
This concurs with the intuition that any contract satisfiable under $\ecc$ or
$\ccc$ is satisfiable under $\scc$ , and any contract that is satisfiable under
$\ecc$ is satisfiable under $\ccc$. Nonetheless, we determine the class of a
contract based on the \emph{weakest} guarantee (among $\ecc$, $\ccc$, and
$\scc$) required to satisfy the contract. The classification scheme is
presented formally in Figure~\ref{sem:classify}; here, $\Delta$ captures
assumptions about the nature of primitive relations, such as $\Rvis$ and $\Rso$
are irreflexive, and the happens-before relation $\hbZ$ is acyclic, preventing
thin-air reads. \KC{Ideally, we need to list all the axioms somewhere.}  The
following definition thus precisely captures our intuition about contract
strength in terms of these rules:

@@

%% The rules to classify contracts into availability classes are straightforward.
%% The remaining rule judges a contract as {\sf WellFormed} if and only if it can
%% be satisfied under strong consistency. The rest are reported as errors.

>>>>

Quantifier-free propositions in our contract language are conjunctions,
disjunctions and implications of predicates expressing relations between pairs
of effect variables. The syntactic class of relations is seeded with primitive
$\visZ$, $\soZ$, and $\sameobjZ$ relations, and also admits derived relations
that are expressible as union, intersection, or transitive
closure\footnote{Strictly speaking, $R^{+}$ is not the transitive closure of
$R$, as transitive closure is not expressible in FOL.  Instead, $R^{+}$ in our
language denotes \emph{a} superset of transitive closure of $R$. Formally,
$R^{+}$ is any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow R'(x,z)$} of
primitive relations.  Commonly used derived relations are the \emph{same object
session order} ($\small \sooZ = \soZ ~\cap~ \sameobjZ$), and the \emph{same
object happens-before order} ($\small \hboZ = (\sooZ ~\cup~ \visZ)^+$).

@@

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}.  Since first-order logic is sound and complete
~\cite{completeness}, $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st}
\Rightarrow \cv_{op}$.  Due to the decidability of our contract language, this
implication check is automatically discharged with the help of a theorem
prover.

@@

\begin{smathpar}
\ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}
\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
	contracts are required to see a causally consistent snapshot of the object
	state, including the actions performed on the same session.  The latter
	requirement entails that if two operations $o_1$ and $o_2$ from the same
	session are applied to two different replicas $r_1$ and $r_2$, the second
	operation cannot be discharged until the effect of $o_1$ is merged with $o_2$
	in both $r_1$ and $r_2$. The \cf{getBalance} operation requires causal
	consistency, as it requires the operations from the same session to be
	visible, which cannot be guaranteed under eventual consistency. We assume
	that causality is only tracked through operations on the same object; two
	operations in the same session but on different objects are considered
	causally unrelated under this definition. Stores typically avoid tracking
	causality across objects to mitigate overheads when causality tracking is
	unnecessary. The corresponding store semantics is captured by the contract
	$\ccc$ presented in Section~\ref{sec:store_sem}.

@@

\noindent Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves
totally ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$.
This concurs with the intuition that any contract satisfiable under $\ecc$ or
$\ccc$ is satisfiable under $\scc$, and any contract that is satisfiable under
$\ecc$ is satisfiable under $\ccc$. Nonetheless, we determine the class of a
contract based on the \emph{weakest} guarantee (among $\ecc$, $\ccc$, and
$\scc$) required to satisfy the contract.

The classification scheme is presented formally in Figure~\ref{sem:classify}.
Along with three straightforward rules that classify contracts into consistency
classes, the classification scheme also presents a rule that judge
well-formedness of a contract. A contract is well-formed if and only if it is
satisfiable under $\scc$ - the strongest possible consistency guarantee any
store can provide. The $\le$ relation used by the rules is defined formally
below:

@@

The $\Delta$ referred in the above defintion captures assumptions about the
nature of primitive relations, such as $\Rvis$ and $\Rso$ are irreflexive, and
the happens-before relation $\hbZ$ is acyclic, preventing thin-air reads.
\KC{Ideally, we need to list all the axioms somewhere.}

### 2014-10-28 core-lang.tex

\section{Operational Semantics}
\label{sec:core-opsem}

We formalize a replicated data store as a triple $\E;\Theta;\Sigma$,
where $\E$ denotes the execution state, $\Theta$ maps replicas to
their local state, and $\Sigma$, the session soup, is the set of
concurrent client sessions interacting with the store.  Operational
semantics given in Figure~\ref{sem:oper} succinctly capture the
behaviour of the store that is relevant to demonstrate the operational
significance of contract classification. For the sake of clarity, we
only consider a single replicated object of well-defined type (for eg:
a replicated object of type \cf{BankAccount}) in our formalization.
Our semantics are parametric over the specification of this replicated
data type.  Figure~\ref{sem:oper} formalizes replicated data type
(RDT) specification as tuple $(\tau,\Ops,\Ctrts)$, where $\tau$ is the
data type, $\Ops$ maps labels ($op$) of operations on $\tau$ to their
definitions, while $\Ctrts$ maps them to their consistency contracts
($\cv$). The definition of an operation is expected to be a lambda
expression in the server-side language\footnote{Since a replicated
object is maintained by a server, we call the language in which
replicated data type operations are defined as \emph{server-side
language}.}, although we do not enforce this in our formalization. We
intentionally leave the server-side language underspecified to
demonstrate that our ideas are independent of the choice of the
programming language used to implement the replicated data type.

@@

A session is modeled simply as a sequence of actions on the replicated
object maintained by the store, where an action is defined as an
application of replicated data type operation to a value. For
technical reasons, we tag every session in the session soup ($\Sigma$)
with a unique {\sf SessID} and the sequence number of next operation
in that session. As with the server-side language, our development is
also orthogonal to the choice of the client-side language, but we
require both the languages to agree on the syntactic class of values
that replicated data type operations ($op$) produce and consume.

An execution state $\E$ is a tuple $\Exec$ where $\EffSoup$ is a set
of effects, and $\visZ,\soZ \subseteq \EffSoup \times
\EffSoup$ are visibility and session order relations over effects,
respectively. An effect ($\eff$), modeled as a tuple $(s,i,op,v)$,
records the fact that $i^{th}$ action in session with {\sf SessID}
$s$, which is an operation $op$ on the replicated object, has been
successfully executed on some replica yielding a return value $v$.
Note that the combination of $s$ and $i$ uniquely identifies the
effect. Session order relation ($\soZ$) relates effects generated by
the same session. An effect $\eff = (s,i,op,v)$ is said to precede
another effect $\eff' = (s',i',op',v')$ in session order if and only
if $s'=s$ and $i'\ge i$. An effect generated at a replica becomes
visible at rest of the replicas eventually. The local state of a
replica $r$ (i.e., $\Theta(r)$) is a set of effects that are currently
visible at $r$. An operation $op$ performed at replica $r$ can only
witness the set of effects ($\Theta(r)$) visible at $r$.  
% Note that in presence of multiple replicated objects, $\visZ$ only
% relates effects of operations performed on a single object. For
% instance, a {\sf withdraw} operation can witness (via $\visZ$) a {\sf
% deposit} operation on the same bank account, but not a {\sf deposit}
% on a different account. 
Often, in our formalization, we use $\visZ$ and
$\soZ$ binary relations to obtain a set of effects visible to a given
effect $\eff$, or set of effects that precede a given effect $\eff$ in
the session order. As a syntactic convenience, whenever $R$ is a
binary relation, we write $R(\eff)$ to denote the set of all $\eff'$
such that $(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$
to denote the set of all $\eff'$ such that $(\eff',\eff) \in R$.

Rule $\rulelabel{Comm}$ captures inter-replica communication. The
basic guarantee provided by the store is causal visibility. That is,
an effect ($\eff$) is made visible at a replica $r$ only after all
the effects that causally precede $\eff$ are made visible at $r$.
Observe that this condition is applicable even for the effects that
originate at replica $r$. 
% It is helpful to think of effects being
% stored in a \emph{write buffer} until causal consistency condition is
% met, at which point they can be merged with the local state
% ($\Theta(r)$). 
It is important to note that that enforcing causal visibility does not
require any inter-replica coordination.

Rule $\rulelabel{Oper}$ is an auxiliary reduction of the
form:\vspace{-1.7mm}
\begin{smathpar}
\auxred{\Theta} {(\E,\langle s,i,op(v) \rangle)} {r} {(\E',\eff)}
\vspace{-1.7mm}
\end{smathpar}
\noindent Under the store configuration $\Theta$, the rule captures
the progress in execution (from $\E$ to $\E'$) due to the application
of operation $op$ to replica $r$ resulting in a new effect $\eff$.
The rule first constructs a \emph{context} for the application from
the local state ($\Theta(r)$) of the replica, by
projecting\footnote{{\textsf{ctxt*}} is auxiliary function
\textsf{ctxt} extended straightforwardly to set of effects} relevant
information from effects in $\Theta(r)$. It then substitutes the
definition ($\Ops(op)$) of the operation for its label ($op$), and
relies on the reduction relation ($\rdtredsto$) of the server-side
language to reduce the application $\Ops(op)(v,~ctxt)$ to a value
$v'$.  Subsequently, the the attributes of execution state, namely
$\EffSoup$, $\visZ$ and $\soZ$, are extended to incorporate
the new effect ($\eff$).

If the operation $op$ is highly-available (rule $\rulelabel{HA}$), we
simply apply the operation to any replica $r$. Since the store
provides causal visibility, highly available operations are
satisfiable under any replica. If the operation is sticky-available
(rule $\rulelabel{SA}$), the operation can be applied to a replica $r$
only if it already contains the effects of all the previous operations
from the same session. This guarantee can be satisfied by applying all
operations from the same session to the same logical copy of the
database. If such a logical copy is untenable, then the operation
might block. Since the store is assumed to converge eventually, the
blocked sticky-available operation guaranteed to unblock eventually.

An unavailable operation expects strong (sequential) consistency.
That is, universe of all effects ($\EffSoup$) in an execution ($\E$)
must be partitionable into a set of effects that \emph{happened
before} $\eff$ and another set that \emph{happened after} $\eff$,
where $\eff$ is the effect generated by an unavailable operation. The
rule $\rulelabel{UA}$ enforces this sequencing in two steps; firstly,
it insists that the the unavailable operation ($op$) witness effects
of all operations executed so far by requiring the global set of
effects $\EffSoup$ to be a subset of local state ($\Theta(r)$) of the
replica ($r$) executing $op$. Secondly, the rule requires the effect
($\eff$) generated by $op$ to be added to the local state of every
other replica in the store, so that further operations on these
replicas can witness the effect of $op$. Since both these steps
require global coordination among replicas, the store is
\emph{unavailable} during the time it is executing $op$. However, it
must be noted that executing unavailable operation does not entail the
convergence of local states of all replicas; it simply means that
there is one replica ($r$) whose local state is the upper bound of
local states of all replicas, and there exists a set containing
atleast one effect ($\eff$), which is their lower bound.

\subsection{Metatheory}

To help us state our main meta-theoretic result, we first define a
well-formedness condition over execution states:

\begin{definition}[Well-formed execution]
An execution state $\E = \Exec$ corresponding to an execution of
operations over a replicated datatype with specification $\rdtspec =
(\tau,\Ops,\Ctrts)$ is said to be well-formed if and only if $\forall
\eff \in \EffSoup. ~\E \models \DDe{[\eff/\cureff]\cv}$, where $\cv =
\Ctrts(\operZ(\eff))$.
\end{definition}

$\models$ denotes the standard \emph{models} relation of
first-order logic.  A valid model for a first-order formula defines
the universe of discourse, and provides interpretations for free
symbols in the formula such that the formula is true under given
interpretations.  An execution state ($\E$) provides interpretations
for $\Rvis$ and $\Rso$ relations, while fixing the universe of
discourse to be the set of all generated effects ($\EffSoup$).
Intuitively, the well-formedness definition captures the idea that for
every generated effect, the contract on the corresponding operation is
satisfied under the given execution state.

Our main meta-theoretic result guarantees the soundness of our
operational semantics in enforcing all contractual obligations by
ensuring that every reduction step preserves the well-formedness of
execution state.

\begin{theorem}[Well-formedness preservation]
\label{lem:core-preservation}
If $\E$ is well-formed and $\E; \Theta; \Sigma \goesto \E'; \Theta';
\Sigma'$, then $\E'$ is well-formed.
\end{theorem}
% Lemmas
%\input{core-lemmas.tex}

The proof of the theorem, which we could not include here owing to
space constraints, can be found in our tech report~\cite{techrep}.

% It should be noted that Theorem ~\ref{lem:core-preservation} only
% proves the safety of our system. Proving its liveness requires us to
% axiomatize eventual consistency.

>>>>

\section{Meta-theory}
\label{sec:core-opsem}

In this section, we present the meta-theoretic result that certifies
the soundness of our classification-based contract enforcement
strategy. 
% We show that any data store can correctly enforce $\scc$,
% $\ccc$, and $\ecc$ guarantees can satisfy application-level
% consistency constraints of all \name contracts.

Fig.~\ref{sem:oper} summarizes the artifacts needed to state our
meta-theoretic result. We formalize an abstract model of a replicated
data store as a tuple $\E;\Sigma$, where $\E$ denotes the execution
state, and $\Sigma$, the session soup, is the set of concurrent client
sessions interacting with the store. The execution state ($\E$) of a
data store is modeled as a tuple $\Exec$ where $\EffSoup$ is a set of
effects during the execution, and $\visZ,\soZ \subseteq \EffSoup
\times \EffSoup$ are visibility and session order relations,
respectively, witnessed over generated effects. A session is modeled
simply as a sequence of replicated data type operations ($op$), tagged
with the consistency class ($\tau$) of their contracts (as determined
by the contract classification scheme). We abbreviate contract classes
($\tau$) as uppercase {\sf SC}, {\sf CC}, and {\sf EC} for brevity.
Their representative contracts ($\cv(\tau)$) are, as usual, written
$\scc$, $\ccc$, and $\ecc$, respectively.  Our contract language and
classification scheme is independent of store semantics. Accordingly,
we abstain from making any specific assumptions about semantics of the
store. However, we do assume the existence of a reduction relation of
form:
\begin{smathpar} 
  \auxred{} {(\E,\Sigma)} {\eff} {(\E',\Sigma')}
\end{smathpar}
\noindent The relation captures store's progress in execution (from
$\E$ to $\E'$) due to the application of an operation from one of the
sessions in $\Sigma$ to a replica, generating a new effect ($\eff$).

Note that an execution state ($\E$) provides interpretations for
primitive relations (eg: $\visZ$) that occur free in contract
formulas, and also fixes the domain of quantification to set of all
effects ($\EffSoup$) observed during the execution. Therefore, $\E$ is
a potential model for any contract formula ($\cv$). If $\E$ is indeed
a model of $\cv$ (i.e., $\E \models \cv$), we say that the store
enforced the contract $\cv$ in execution $\E$.

@@

We now state the soundness of contract enforcement as following:

\begin{theorem}[Soundness of Contract Enforcement]
\label{lem:core-preservation}
Let $\cv$ be the contract of a replicated data type operation $op$,
and let $\tau$ denote the consistency class of $\cv$ 
as determined by the contract classification scheme. Forall execution
states $\E$, $\E'$ such that 
$\auxred{} {(\E,\langle (op,\tau);\,\sigma \rangle \pll \Sigma)} {\eff} 
 {(\E',\langle \sigma \rangle \pll \Sigma)}$
if $\E' \models [\eff/\cureff]\,\cv(\tau)$, then $\E' \models [\eff/\cureff]\,\cv$ 
\end{theorem}

The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ guarantees in all executions, then contract
classification scheme can be used to satisfy any \name contract on
that data store. The proof of the theorem has been included in our
tech report~\cite{techrep}. Tech report also contains operational
semantics of our implementation of the data store, which concretizes
the reduction relation ($\xhookrightarrow{}$), along with the proof
that it correctly enforces all \name contracts.
